---
title: "BMI 510: Homework #2"
author: "Elizabeth Nemeti"
date: "01/28/2023"
output:
  html_document: default
  pdf_document: default
---

***
This homework is due Monday, 1/23 at 1 PM. (It may be returned up until Wednesday, 1/25 at 1 PM without penalty, but the next homework will be released 1/23.)
***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

# Q1

Let E and F be two events for which P(E or F) = 0.75. What is the probability that neither E nor F occurs?

```{r }
# answer code for Q1
1 - 0.75 = 0.25
```

The probability that neither E nor F occurs is 0.25.

# Q2
Let’s say you have an experiment with two possible outcomes. One outcome has probability p, and the other outcome has probability p^2.


```{r}
# answer code for Q2a
p=(-1+sqrt(5))/2
p

# answer code for Q2b
1 + p

```

You need to use the quadratic formula and p is 0.618...

# Q3

A coin is tossed three times and the sequence of heads and tails is recorded. List the sample space.

```{r }
N/A
```

S = {HHH,HHT,HTH,THH,HTT,THT,TTH,TTT}

# Q4

With the same coin, list the elements that make up the following events A-C.
A. At least two tosses are heads.
B. The first two tosses are heads.
C. The last toss is a tail.

```{r }
N/A
```

A = {HHH,HHT,HTH,THH}
B = {HHH,HHT}
C = {HHT,HTT,TTT,THT}

# Q5

Write a function to convert probabilities to odds, and a function to convert odds to probabilities.

```{r }

probability_to_odds <- function(probability) {
  return(probability / (1 - probability))
}

odds <- probability_to_odds(0.5)
odds

odds_to_probability <- function(odds) {
  return(odds / (odds + 1))
}

probability <- odds_to_probability(1)
probability
```

The odds are 1:1 when you pass in 0.5 for the 1st function.
The probability is 0.5 when you pass in 1 for the 2nd function.

# Q6

A) Graph odds O(A) as a function of probability P(A) as P(A) varies between 0 and 1.
B) Graph probability P(A) as a function of odds O(A) as O(A) varies between 0 and 100.

```{r }
library(ggplot2)

# A)
probability <- seq(0, 1, by = 0.01)
odds <- probability_to_odds(probability)
data <- data.frame(probability, odds)

ggplot(data, aes(x = probability, y = odds)) +
  geom_line() +
  ggtitle("Odds as a function of probability") +
  xlab("Probability") +
  ylab("Odds")

# B)
odds <- seq(0, 100, by = 1)
probability <- odds_to_probability(odds)
data <- data.frame(odds, probability)

ggplot(data, aes(x = odds, y = probability)) +
  geom_line() +
  ggtitle("Probability as a function of odds.") +
  xlab("Probability") +
  ylab("Odds")

```

By using ggplot2, I made the two graphs where graph B is the inverse of graph A. 

# Q7

There are a lot of visualization softwares for DAGs, but ASCII art works fine for simple ones (e.g., X1 --> X2 ). Draw the DAGs representing the following joint probability distributions.

```{r }
N/A
```

1. X1 X2 (no arrow)
2. X1 --> X2
3. X2 --> X1
4. X1 --> X2 --> X3
5. SES --> water --> cholera

6.     --> water --> cholera
   SES 
       --> elevation


# Q8
Because the uniform distribution on [0,1] has mean 1/2 and variance 1/12, the sum of 12 random variables drawn from the uniform distribution (minus 6) has mean 0 and variance 1. Use a computer to simulate multiple samples of 12 uniform random variables, and sum each sample. Use base R to draw a histogram of the sums; then do the same thing with ggplot2.


```{r }
sums <- rowSums(matrix(runif(12*1000), ncol = 12)) - 6

#without ggplot
hist(sums, main = "Histogram of Sums", xlab = "Sum")

#with GG plot
library(ggplot2)
ggplot(data = data.frame(sums), aes(sums)) + 
  geom_histogram(fill = "grey",bins = 20) + 
  labs(title = "Histogram of Sums", x = "Sum", y = "Frequency")

```

# Q9

Write a short program that calculates a single random variable drawn from N(0,1) based on the answer to problem 8.

```{r }
random_vars <- runif(12)
sum_of_vars <- sum(random_vars) - 6
scaled_var <- sum_of_vars / sqrt(12)
single_var <- scaled_var
single_var

```

# Q10
The data we used to demonstrate “Simpson’s paradox” are available at
[https://jlucasmckay.bmi.emory.edu/global/bmi585/simpson_data.csv. Use ggplot2 to approximate the graph from the lecture.

```{r }
simpson_data <-read.csv("simpson_data.csv", header=TRUE, sep=",")
View(simpson_data)

ggplot(simpson_data, aes(x = Exercise, y = Cholesterol, color = factor(Age))) + 
  geom_point(size=0.5, shape=20) + #fixing point sizes
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) + #fixing axis range
  geom_smooth(method = "lm", se = FALSE) + # here we add the regression lines/ fit per group
  xlab("Exercise") +
  ylab("Cholesterol") +
  ggtitle("(Bad) Cholesterol vs. Exercise") +
  scale_color_discrete(name = "Age") + # here we add the colour scale to the Age factor
  theme_classic() + # setting a white background
  theme(
        axis.title = element_text(size=5),
        plot.background = element_rect(fill = "white"),
        panel.grid.major = element_line(color = "lightgrey"), # grey gridlines
        panel.grid.minor = element_line(color = "lightgrey"),
        axis.text.x = element_text(size = 5), # resizing axes
        axis.text.y = element_text(size = 5),
        legend.title = element_text(size = 5), # resizing legend
        legend.text = element_text(size = 3), 
        legend.position = c(0.05, 0.79) # moving the legend from right to custom spot
        )

```

