---
title: "Homework 5"
author: "J. Lucas McKay"
date: "2/13/2023"
output: html_document
---

<style>div.soln { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}</style>

```{r setup, include=FALSE, echo=T}
knitr::opts_chunk$set(echo = F, message = F, warning = F, collapse = T)
library(tidyverse)
```

***
1. Using only `rnorm`, draw a sample of 1000 random numbers from $\chi^2_7$. Plot a histogram. **1 point**

```{r setup, include=FALSE}

rand <- matrix(rnorm(7 * 1000), ncol = 7) # 7 independent standard normal random variables
sample <- rowSums(rand^2) # obtain sample from chi-squared distribution
hist(sample, breaks = 30, col = "blue", xlab = "Value", main = expression(paste(chi[7]^2, " Distribution")))

```

***
2. Plot the log-likelihood of the sample assuming a $\chi^2_{df}$ distribution under candidate `df` values from 1 to 14. **1 point**

```{r setup, include=FALSE}

# create function to calculate a log-likelihood
df <- 14

log_likelihood_func <- function(df, sample) {
  n <- length(sample)
  k <- df/2
  log_likelihood_singular <- -(n*k)*log(2) - n*lgamma(k) + (k-1)*sum(log(sample)) - (1/2)*sum(sample)
  return(log_likelihood_singular)
}

# calculating the log-likelihoods for degrees of freedom ranging from 1 to 14
df_vals <- seq(1, 14)
log_likelihood <- sapply(df_vals, log_likelihood_func, sample = sample) 

#plotting histogram
plot(df_vals, log_likelihood, type = "l", xlab = "Degrees of Freedom", ylab = "Log-Likelihood", main = "Log-Likelihood for Chi-Squared Distribution")

```

***
3. Using only `rnorm` (and potentially your answer from question 1) draw a sample of 1000 random numbers from $t_7$. Plot a histogram. **1 point**

```{r setup, include=FALSE}

# setting sample size
n <- 1000 

# Generate a sample of 1000 random numbers from a standard normal distribution using rnorm
rand_nums <- rnorm(n)

# generate random numbers from a chi-squared distributionn
rchi <- function(df) {
  sum(rnorm(df)^2)
}

# sample of random numbers from the chi-squared distribution
rand_chi_nums <- replicate(n, rchi(7))

# generate sample of 1000 random numbers from a t-distribution 
t_dist <- rand_nums / sqrt(rand_chi_nums/7)

# plotting histogram
hist(t_dist, breaks = 30, col = "blue", xlab = "T distibution", main = expression(paste("Histogram of ", t_dist[7], " Random Sample")))

```

***
4. Plot the log-likelihood of your sample assuming a $t_{df}$ distribution under candidate `df` values from 1 to 14. **1 point**

```{r setup, include=FALSE}

df <- seq(1, 14)
sample <- 1000

log_likelihood_func <- function(df, sample) {
  n <- length(sample)
  k <- df
  log_likelihood_singular <- -(n*(k+1)/2)*log(1+(1/k)*sum(sample^2)) + (-(n/2))*log(pi) - lgamma((k+1)/2) + ((k+1)/2)*sum(log(1 + (1/k)*sample^2))
  return(log_likelihood_singular)
}

# calculating the log-likelihoods for degrees of freedom ranging from 1 to 14
log_likelihood <- sapply(df, log_likelihood_func, sample = sample)

#plotting histogram
plot(df, log_likelihood, type = "l", xlab = "Degrees of Freedom", ylab = "Log-Likelihood", main = "Log-Likelihood for t-distribution")

```
***
5. Using only `rnorm` (or potentially your answer from question 1), draw a sample of 1000 random numbers from $F_{4,4}$. Plot a histogram. **1 point**. 

```{r setup, include=FALSE}

#using only rnorm
set1 <- rnorm(1000, mean = 0, sd = 1)
set2 <- rnorm(1000, mean = 0, sd = 1)
set3 <- rnorm(1000, mean = 0, sd = 1)
set4 <- rnorm(1000, mean = 0, sd = 1)

# Compute the F ratio
f_ratio <- ((set1^2 + set2^2)/(2)) / ((set3^2 + set4^2)/(2))

# Plot the histogram
# breaks at 20 so you see frequency up to 1000

hist(f_ratio, breaks = 20, col = "blue", xlab = "Sample Values", main = "Sample from F-distribution")

```

***
6. Plot the log-likelihood of your sample assuming a $F_{df_1,df_2}$ distribution with $df_1$ fixed at 4 under candidate `df_2` values from 1 to 14. **1 point**

```{r setup, include=FALSE}

df1 <- 4
df2 <- seq(1, 14)
sample <- 1000

log_likelihood_func <- function(df2, sample) {
  n <- length(sample)
  k1 <- df1/2
  k2 <- df2/2
  log_likelihood_singular <- -(n*(k1+k2))*log(2) - n*lgamma(k1) - n*lgamma(k2) + (k1-1)*sum(log(sample)) + (k2-1)*sum(log(1+((k1*sample)/(k2)))) - ((k1+k2)*sum(log(1+((sample/k2))))) - (k1*log(k1)) - (k2*log(k2))
  return(log_likelihood_singular)
}

log_likelihood <- sapply(df2, log_likelihood_func, sample = sample)

plot(df2, log_likelihood, type = "l", xlab = "df_2", ylab = "Log-Likelihood", main = "Log-Likelihood for F(4,df_2) distribution")


```

***
7. Write a short function to convert a test statistic $x$ into the area under the $z$-density for values $≥|x|$. This area is the P value. The function should take $x$ as an argument. **(1 point)**

```{r setup, include=FALSE}

# function to convert the test statistic into P value
pvalue_convert <- function(x) {
  p <- 1 - pnorm(abs(x))
  return(p)
}

```


***
8. Write a short function to convert a test statistic $x$ into the area under the $t_{df}$-density for values $≥|x|$. This area is the P value. The function should take $x$ and $df$ as arguments. **(1 point)**

```{r setup, include=FALSE}

tdist_pvalue_convert <- function(x, df) {
  p <- 1 - pt(abs(x), df) # where pt func is for cumulative dist for t dist
  return(p)
}

```

***
9. Simulate the results of a cancer trial.

Let's say that in a double-blind randomized clinical breast cancer trial, the primary outcome is *tumor diameter*.
At study end, the average tumor reduction in tumor size was 3 ± 1 cm (mean ± sd) in the treatment group and 0 ± 1 cm (mean ± sd) in the control group.
Simulate reduction values for N = 12 treatment and N = 12 control participants. Use `t.test` to whether the reductions differ between groups.
Compare the results when using the default options and when using `var.equal = T`. **(1 point)**

```{r setup, include=FALSE}

# set seed to get the same values for Q10 comparison
set.seed(42)

# Simulate reduction values
treatment_group <- 12
control_group <- 12

# average tumor reduction in tumor size was 3 ± 1 cm (mean ± sd)
treatment_reductions <- rnorm(treatment_group, mean = 3, sd = 1)
# 0 ± 1 cm (mean ± sd) in the control group
control_reductions <- rnorm(control_group, mean = 0, sd = 1)

# Use `t.test` for whether the reductions differ between groups
# using default options
reduction_vals <- t.test(treatment_reductions, control_reductions)
reduction_vals

# Compare the results when using the default options and when using `var.equal = T`
reduction_vals_varequal <- t.test(treatment_reductions, control_reductions, var.equal = T)
reduction_vals_varequal

```

Default where Var.equal is not true

Welch Two Sample t-test
data:  treatment_reductions and control_reductions
t = 7.2188, df = 19.469, p-value = 6.442e-07
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval: 2.164074 3.927354
sample estimates:
mean of x  mean of y 
2.9902808 -0.0554334 

Var.equal = True

Two Sample t-test
data:  treatment_reductions and control_reductions
t = 7.2188, df = 22, p-value = 3.111e-07
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 2.170717 3.920711
sample estimates:
 mean of x  mean of y 
 2.9902808 -0.0554334 
 
The variances of the treatment and control groups were not equal, so using the default option where var.equal = F is the appropriate choice, as the var.equal = True option induces error. 

***
10. Recreate the functionality of `t.test`. Use base R functions (e.g., `length`, `mean`, `var`, `sqrt`) and `pt` to recreate the t statistic and p value from problem 9 without `t.test` using `var.equal=TRUE`. **(1 point)**

```{r setup, include=FALSE}
# set seed to get the same values as Q9 
set.seed(42)

# using variables from Q9
treatment_group <- 12
control_group <- 12
treatment_reductions <- rnorm(treatment_group, mean = 3, sd = 1)
control_reductions <- rnorm(control_group, mean = 0, sd = 1)

# recreating t test and p value
t_statistic <- (mean(treatment_reductions) - mean(control_reductions)) / sqrt((var(treatment_reductions)/treatment_group) + (var(control_reductions)/control_group))

pvalue <- 2 * pt(-abs(t_statistic), df = treatment_group + control_group - 2)

cat("The t statistic is:", t_statistic)
cat("The p-value is:", pvalue)

```

This results in the same t statistic and p value as in Q9 where var.equal = True. 

Since rnorm is generating a different set of numbers each time, I set the seed so you can see that Q10 does actually reproduce t test.

