---
title: "Homework 6"
author: "J. Lucas McKay"
date: "2/20/2023"
output:
  pdf_document: default
  html_document: default
student: Elizabeth Nemeti
---

<style>div.soln { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}</style>

***
```{r setup, include=FALSE}
setwd("/Users/elizabethnemetichipkes/Desktop/R for BMI510")
library(tidyverse)
library(pwr)
library(ggplot2)
```

Section 1: Write a wrapper function around power.t.test that maps power to minimum
sample size.

```{r Secion 1, include=TRUE}

# Q1
min_samplesize_func <- function(d, power) {
  power.t.test(d = d, power = power)$n
}

#Q2
# define power levels and effect sizes aka d
power_levels <- c(0.8, 0.9, 0.95)
d <- seq(0.2, 2.0, 0.001)

df <- expand.grid(d = d, power = power_levels)
df$n <- mapply(min_samplesize_func, df$d, df$power)

# create linear plot
linear_plot <- ggplot(df, aes(x = d, y = n, color = factor(power))) +
  geom_line() +
  scale_color_manual(values = c("#330000", "red", "#CC9933")) +
  scale_x_continuous(limits = c(0, 2), breaks = seq(0, 2, 0.5)) +
  scale_y_continuous(limits = c(0, 600), breaks = seq(0, 600, 200)) +
  xlab("d") +
  ylab("n") +
  theme(panel.background = element_rect(fill = "white"), 
        panel.grid.major = element_line(color = "darkgrey"), 
        panel.grid.minor = element_line(color = "darkgrey"),
        legend.position = "bottom",
        legend.title = element_blank())

linear_plot

d <- seq(0.2, 2.0, 0.5)
# create log plot
log_plot <- ggplot(df, aes(x = d, y = n, color = factor(power))) +
  geom_line() +
  scale_color_manual(values = c("#330000", "red", "#CC9933")) +
  scale_x_continuous(limits = c(0, 2), breaks = seq(0, 2, 0.5)) +
  scale_y_log10(limits = c(5,800), breaks = c(10, 30, 100, 300)) +
  xlab("d") +
  ylab("n") +
  theme(panel.background = element_rect(fill = "white"), 
        panel.grid.major = element_line(color = "darkgrey"), 
        panel.grid.minor = element_line(color = "darkgrey"),
        legend.position = "bottom",
        legend.title = element_blank())

log_plot

```

Why does the logarithmic plot have a “staircase” shape for high values of d?

Since this plot has a logarithmic y axis, the staircase pattern emerges when we see the value of d increasing and n (y value) increasing at a much larger rate than d. The stairs come from small differences in the y axis that are much larger than the changes in the x axis. 

Section 2: Fisher’s exact test.

```{r Secion 2, include=TRUE}

m <- 35  #promotions
n <- 13  #failures
t <- 24  #total (not 48 bc drawn from left column)
x <- 21  #top left cell

prob_promote_male <- dhyper(x, m, n, t)
prob_promote_male

#create a matrix
cont_table <- matrix(c(21, 3, 24, 14, 10, 24, 35, 13, 48), nrow = 3, ncol = 3, byrow = TRUE)
rownames(cont_table) <- c("promote", "hold", "total")
colnames(cont_table) <- c("male", "female", "total")
# View(cont_table)

# Calculate probabilities for all cells
prob_hold_male <- dhyper(3, 13, 35, 24) 
prob_promote_female <- dhyper(14, 35, 13, 24)
prob_hold_female <- dhyper(10, 13, 35, 24)

# Chisq test
chisq <- chisq.test(cont_table)
p_value <- chisq$p.value

```

2.05% is for the top left cell. 

Since the degrees of freedom for Fisher's exact test equals r-1 x c-1, here the contingency table has 3 rows and 3 columns, so it's 4. We need to test 4 cells to perform Fisher's exact test.

P value was 0.27, which would not reject the null hypothesis, meaning there could be other factors involved in women not getting as promoted as men in this case. 

Section 3: Confidence Interval Coverage


```{r Secion 3, include=TRUE}

# Write a function waldci(p_hat,n,alpha) that returns the upper and lower bounds of a Wald confidence interval for the probability of success for a sample of size with observed successes.
waldci <- function(p_hat, n, alpha) {
  z_value <- qnorm(1 - alpha/2)
  upper_bound <- p_hat + z_value*sqrt(p_hat*(1-p_hat)/n)
  lower_bound <- p_hat - z_value*sqrt(p_hat*(1-p_hat)/n)
  return(c(lower_bound, upper_bound))
}

# Write a function coverageProb(p,n_trials,n_replicates,alpha) that simulates n_replicates Bernoulli trials of n_trials each with probability of success p , calculates a Wald confidence interval for each, and returns the proportion of the confidence intervals that actually contain the true underlying success probability p

coverageProb <- function(p, n_trials, n_replicates, alpha) {
  simulate_bern_trials <- replicate(n_replicates, rbinom(n_trials, 1, p))
  
  p_hat <- apply(simulate_bern_trials, 2, mean)
  standard_error <- sqrt(p_hat*(1-p_hat)/n_trials)
  
  waldci_vals <- waldci(p_hat, n_trials, alpha)
  
  z_value <- qnorm(1 - alpha/2)
  lower_bound <- p_hat - z_value*standard_error
  upper_bound <- p_hat + z_value*standard_error
  
  coverage <- sapply(1:n_replicates, function(i) {
    p >= lower_bound[i] & p <= upper_bound[i]
    # p >= waldci_vals[1] & p <= waldci_vals[2]
  })
  
  proportion_coverage <- sum(coverage)/n_replicates
  return(proportion_coverage)
}

p_values <- seq(0.001, 0.999, by = 0.001)
n_trials <- 50
n_replicates <- 100
alpha <- 0.05

sample12 = c()
sample50 = c()
sample1000 = c()

for (val in p_values){
  sample12 <- append(sample12, coverageProb(val, 12, n_replicates, alpha))
  sample50 <- append(sample50, coverageProb(val, 50, n_replicates, alpha))
  sample1000 <- append(sample1000, coverageProb(val, 1000, n_replicates, alpha))

  }

library(ggplot2)
df <- data.frame(p_values, sample12, sample50, sample1000)
colors <- c("12" = "darkorange", "50" = "salmon", "1000" = "red")
wald_ci_plot <- ggplot(df, aes(x = p_values)) +
  geom_line(aes(y = sample12, color = "12"), size = 0.5) + 
  geom_line(aes(y = sample50, color="50"), size = 0.5) +
  geom_line(aes(y = sample1000, color="1000"), size = 0.5) +
  labs(x = "p",
        y = "Probability that Wald CI covers p",
        color = "Sample size") +
  # scale_color_manual(values = c("darkorange", "salmon", "red")) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.5)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.5)) +
  theme(panel.background = element_rect(fill = "white"), 
        panel.grid.major = element_line(color = "darkgrey"), 
        panel.grid.minor = element_line(color = "darkgrey"),
        legend.position = "right")+
  scale_color_manual(values = colors)

wald_ci_plot

```



