---
title: "Homework 7"
author: "J. Lucas McKay"
date: "3/17/2023"
output:
  pdf_document: default
  html_document: default
student: Elizabeth Nemeti
---

<style>div.soln { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}</style>

***
```{r setup, include=FALSE}
setwd("/Users/Elizabeth Nemeti/Desktop/R for BMI510")
library(tidyverse)
```

Section 1: Multiple Comparisons

We discussed several methods for multiple comparison adjustment, all of which can be implemented from scratch. In particular, the built-in p.adjust function returns “adjusted” p-values, which may not be what you want as an analyst. One way to improve the utility is to create “wrapper” functions around p.adjust.

Implement a Bonferroni-Holm correction function that takes a vector of p-values p and a value for alpha and returns a logical vector indicating which elements of p are statistically significant. Construct this as a“wrapper” around the built-in p.adjust function.Create a similar function for FDR.

Test your function on the following vector of p-values.
NB: be sure that the functions work with unordered inputs!

## Test A Test B Test C Test D Test E Test F Test G Test H Test I Test J ## 0.0025 0.0050 0.0075 0.0100 0.0125 0.0150 0.0175 0.0200 0.0225 0.0250

```{r Secion 1.1, include=TRUE}

p <- c(0.0025, 0.0050, 0.0075, 0.0100, 0.0125, 0.0150, 0.0175, 0.0200, 0.0225, 0.0250)

alpha <- 0.05

bonferroni_holm_correction <- function(p, alpha) {
  p_adjusted <- p.adjust(p, method = "holm")
  sig_element <- p_adjusted <= alpha
  return(sig_element)
}

print(bonferroni_holm_correction(p, alpha))

fdr_correction <- function(p, alpha) {
  p_adjusted <- p.adjust(p, method = "fdr")
  sig_element <- p_adjusted <= alpha
  return(sig_element)
}

print(fdr_correction(p, alpha))

```

Revise the function from part 1 so that you calculate adjusted p-values manually.

```{r Secion 1.2, include=TRUE}

bonferroni_holm_correction_manual <- function(p, alpha) {
  n <- length(p)
  p_sorted <- sort(p)
  rank <- order(p)
  p_adjusted <- pmin(1, n * p_sorted / (n:1))
  p_adjusted <- p_adjusted[order(rank)]
  sig_element <- p_adjusted <= alpha
  return(sig_element)
}

print(bonferroni_holm_correction_manual(p, alpha))

```

Revise the function from part 2 so that you calculate adjusted p-values manually.

```{r Secion 1.3, include=TRUE}

fdr_correction_manual <- function(p, alpha) {
  n <- length(p)
  p_sorted <- sort(p)
  rank <- order(p)
  p_adjusted <- pmin(1, p_sorted * n / (1:n))
  p_adjusted <- p_adjusted[order(rank)]
  sig_element <- p_adjusted <= alpha
  return(sig_element)
}

print(fdr_correction_manual(p, alpha))

```
Section 2

There is a set of CSF biomarker data available here (https://jlucasmckay.bmi.emory.edu/global/bmi510/csf.csv).Implement an ANOVA on the variable Ab42 using the grouping variable Group. You should perform a “complete case” analysis; e.g., delete any observations with missing values for Ab42.

```{r Secion 2.1, include=TRUE}

csf_data <- read.csv("csf.csv", header=TRUE, sep=",")
View(csf_data)

# remove rows with Nans for Ab42
csf_data <- csf_data[!is.na(csf_data$Ab42),] 
View(csf_data)

```

Calculate SSGroup.(1 point)
(sum of squares between groups)

```{r Secion 2.2, include=TRUE}

group_means <- tapply(csf_data$Ab42, csf_data$Group, mean)
SSGroup <- sum((group_means-mean(csf_data$Ab42))^2*table(csf_data$Group))
print(SSGroup)

```
Calculate SStot.(1 point)
(total sum of squares)

```{r Secion 2.3, include=TRUE}

SSTot <- sum((csf_data$Ab42-mean(csf_data$Ab42))^2)
print(SSTot)

```
Calculate SSerror.(1 point)
(residual sum of squares)

```{r Secion 2.4, include=TRUE}

SSError <- SSTot-SSGroup
print(SSError)

```
Calculate MSGroup.(1 point)
(mean square between groups)

```{r Secion 2.5, include=TRUE}

MSGroup <- SSGroup/(length(unique(csf_data$Group))-1)
print(MSGroup)

```
Calculate MSError.(1 point)
(mean square within groups)

```{r Secion 2.6, include=TRUE}

MSError <- SSError/(nrow(csf_data)-length(unique(csf_data$Group)))
print(MSError)

```
Calculate the F-statistic.(1 point)

```{r Secion 2.7, include=TRUE}

F_statistic <- MSGroup/MSError
print(F_statistic)

```
Calculate the p-value(1/2 point)and check it against the result of anova.(1/2 point)

```{r Secion 2.8, include=TRUE}

p_value <- 1-pf(F_statistic, (length(unique(csf_data$Group))-1), (nrow(csf_data) - length(unique(csf_data$Group))))
print(p_value)

summary(aov(Ab42~Group, data=csf_data))

```

